% R Notes. Package data.table
% arostaing@outlook.com
% March, 2014

---

- Speed improvement
- Extends and enhances the functionality of data.frame
- Different syntax from data.frame

---


## Sample data

R and some packages come with sample data included.

```{r sampleData1, results='hide'}
require(ggplot2)
# Use data() to discover available data sets in each package
data()
```

I'll use **movies** dataframe included in **ggplot2** package.  
You can see the data's help file typing `? movies`.   
The sample is a data frame with 28819 rows and 24 variables.  

The columns are:  

- *title*. Title of the movie.  
- *year*. Year of release.  
- *budget*. Total budget (if known) in US dollars.  
- *length*. Length in minutes.  
- *rating*. Average IMDB user rating.  
- *votes*. Number of IMDB users who rated this movie. 
- *r1-10*. Multiplying by ten gives percentile (to nearest 10%) of users who rated this movie a 1.
- *mpaa*. MPAA rating.  
- *action, animation, comedy, drama, documentary, romance, short*. Binary variables representing if movie was classified as belonging to that genre.

Load sample data:
```{r sampleData2}
data(movies)
head(movies, 5)
```

You can see *data.table* use samples with `example` command.
```{r sampleData3, results='hide', echo=TRUE}
require(data.table)
example(data.table)
```

---

## Creating data.table

Creating *data.table* is like creating data.frames.
```{r Creating1, results='hold', echo=TRUE}
require(data.table)
DF = data.frame(x=rep(c("a","b","c"),each=3), y=c(1,3,6), v=1:9)
DT = data.table(x=rep(c("a","b","c"),each=3), y=c(1,3,6), v=1:9)
DF
DT
```
By default *data.frame* turns characters data into factors, **while data.table does not**
```{r Creating2, results='hold', echo=TRUE}
str(DF)
str(DT)
```

It is also possible to create a *data.table* out of an existing *data.frame*.
```{r Creating3, results='hold', echo=TRUE}
DF2 <- data.table(DF)
DF2
rm(DF2)
```

Lets load *movies* *data.frame* into a *data.table*
```{r Creating4, results='hold', echo=TRUE}
# Create data.table
moviesdt <- data.table(movies)

# Printing data.table just prints the first five and last five rows.
moviesdt
```

To see some information about *data.table* in memory:
```{r Creating5, results='hold', echo=TRUE}
tables()
```

---

## Accesing data

Accessing rows is similar to accessing *data.frame*'s row.
```{r Accessing1, results='hold', echo=TRUE}
# Rows 1 - 5
moviesdt[1:5,]
```
```{r Accessing2, results='hold', echo=TRUE}
# Year == 2005 & rating > 9
moviesdt[moviesdt$year == 2005 & movies$rating > 9.5,]
```

Accesing columns is quite different, with *data.frames* columns should be specified as a character vector.  
With *data.tables* the columns should be specified as a list of the actual names, not as characters.
```{r Accessing3, results='hold', echo=TRUE}
# Sample subset...
sampledt <- head (moviesdt[movies$rating > 9.5,], 10)

# Access columns year and title
sampledt[, list(year, title)]
```

It's also possible specify the columns names as a character, but in this case the *with* argument should be set to *FALSE*.
```{r Accessing4, results='hold', echo=TRUE}
# Access columns year and title
sampledt[, c('year', 'title'), with=FALSE]
```


