% R Notes. Aggregation
% arostaing@outlook.com
% March, 2014

---

## Sample data

R and some packages include some sample data. 

```{r sampleData1, results='hide'}
# Use data() to discover available data sets in each package
data()
```

I'll use **mtcars** dataframe.  
You can see the data's help file typing `? mtcars`.   
The sample is a data frame with 32 observations on 11 variables.   
The Format is: 

- [, 1]  mpg         Miles/(US) gallon
- [, 2]	 cyl	 Number of cylinders
- [, 3]	 disp	 Displacement (cu.in.)
- [, 4]	 hp	 Gross horsepower
- [, 5]	 drat	 Rear axle ratio
- [, 6]	 wt	 Weight (lb/1000)
- [, 7]	 qsec	 1/4 mile time
- [, 8]	 vs	 V/S
- [, 9]	 am	 Transmission (0 = automatic, 1 = manual)
- [,10]	 gear	 Number of forward gears
- [,11]	 carb	 Number of carburetors

Load sample data:
```{r sampleData2}
data(mtcars)
str(mtcars)
```

---

## Aggregation

### data.frame aggregation

Formulas: 

- Left side represents a variable that we want to make a calculation on 
- Right side represents one or more variable that we want to group the calculation by.

```{r data.frame.aggregation1}
# Lets see car's horsepower mean by number of cylinders
result <- aggregate (hp ~ cyl, mtcars, mean)
result[order(result$hp, decreasing=TRUE),]
```

To group the data by more than one variable, add the additional variable to the right side of the formula separating it with a plus sign.

```{r data.frame.aggregation2}
# Lets see car's horsepower mean by number of cylinders and number of carburetours
result <- aggregate (hp ~ cyl + carb, mtcars, mean)
result[order(result$hp, decreasing=TRUE),]
```

To aggregate more than one variable, they must be combined using *cbind* on the left side of the formula.
```{r data.frame.aggregation3}
# Add 1/4 mile time mean
result <- aggregate (cbind(hp, qsec) ~ cyl + carb, mtcars, mean)
result[order(result$hp, decreasing=TRUE),]
```

Do you need using more than one function? `plyr` package helper function make it easy.  
`each`: Combine multiple functions into a single function returning a named vector of outputs.
```{r data.frame.aggregation4}
require(plyr)
result <- aggregate (cbind(hp, qsec) ~ cyl + carb, mtcars, each(min,max,mean))
result[order(result$cyl, result$carb, decreasing=TRUE),]
```
The result is a named vector of outputs.
```{r data.frame.aggregation5}
str (result)
```

In Our sample, there are two named vectors
```{r data.frame.aggregation6}
# Vector with values for results of functions min,max and mean of the hp column
class(result$hp)
result$hp
# Access min results...
result$hp[,1]
```

```{r data.frame.aggregation7}
# Vector with values for results of functions min,max and mean of the hp column
class(result$qsec)
result$qsec
# Access qsec results...
result$qsec[,1]
```

---
